---
title: "REU_2017"
author: "Wesner"
date: "January 17, 2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## REU_2017 Analysis

##Packages

```{r message=FALSE,warning=FALSE}
library(tidyverse)
library(brms)
library(ggridges)
library(viridis)
library(RCurl)
```

##Diet data
In the console, make the diet_abund dataframe a .RData file\
*save(diet_abund, file="diet_abund.RData")*

```{r, warning=FALSE}
x <- getURL("https://raw.githubusercontent.com/jswesner/reu_bcra/master/data_abund.csv")
diet_abund <- read.csv(text = x)

diet_abund2 <- diet_abund %>%
  select(c(-chiro_tot,-total)) %>%
  mutate(total = rowSums(.[8:26]),
         date = as.Date(date, format = '%m/%d/%Y'))
head(diet_abund2)
diet_abund2$method <- ifelse(diet_abund2$method=="dissection","dissection", "lavage")
```


#Relationship between fish size and number of prey in guts
Larger bluegills appear to have more prey items, but no other species show this size-number relationship
```{r,echo=FALSE,warning=FALSE,message=FALSE}
ggplot(subset(diet_abund2,method!="field_lavage + dissection"), aes(x=length_mm,y=total,color=method))+
  geom_point()+
  facet_grid(species~date)+
  geom_smooth(method="lm")+
  NULL


````


#Compare results from gastric lavage and gut dissections
Dissections have more prey (they should), but most dissections also occurred on the first date, so it is\ impossible to know how much of the difference is due to date versus method. 

```{r, echo=FALSE,warning=FALSE}
ggplot(subset(diet_abund2,method!="field_lavage + dissection"&species!="quillback"), aes(x=species,y=total,group=as.factor(date),color=as.factor(date),shape=method))+
  geom_boxplot(aes(group=interaction(date,species)),position=position_dodge(width=.75))+
  geom_point(position=position_dodge(width=.75),size=2)+
  #facet_grid(species~date)+
  #geom_smooth(method="lm")+
  scale_color_grey()+
  scale_y_log10()+
  NULL
```

We quantified these differences using a Bayesian generalized linear model with a Poisson likelihood (since\ the number of diet items (response variable) was positive and integer. The models below measure gut fullness as a function of species, date, and their interaction. 

#Choose the priors
To ensure that the model reflected reasonable prior beliefs, we plotted the prior predictive distribution under several\ iterations of prior specificiations. Starting with a prior for the intercept at N(3,2) and betas at N(0,5). The chunk\ below specifies the model and code to make plots. Note the line in the *brm()* statment that says\ *sample_prior="only"*.\ This statement tells the model so ignore the likelihood when fitting the model. In otherwords,\ there is no data included in this model. The data statement in the code is there simply as a placeholder. It tells the\ *brm* model how to structure the output (i.e. we want the interactions between species and date2 with names included in\ the data frame from *diet_abund2*, but we don't want the data itself yet).\ 

Here's the code.
```{r cache = T, warning = F, message = F, results = "hide"}
diet_abund2$total_1 <- diet_abund2$total + 1 #add 0.01 to remove zeros (poisson can't handle zeros)
diet_abund2$date2 <- as.factor(diet_abund2$date) #make date a factor

model_diet_prior1 <- brm(total ~ species*date2, data=subset(diet_abund2,species!="quillback"),
                         family=poisson(link="log"),
                         prior=c(prior(normal(3,2),class="Intercept"),
                         prior(normal(0,5),class="b")),
                         chains=4, iter=1000,sample_prior="only")  


``` 
Now let's plot the posteriors to see what the priors imply. The first step is to extract posteriors of the marginal distribution for each treatment combination we're interested in. First, extract posterior samples from the HMC chains. We do this using the *marginal_effects* function in *brms*, which samples from the posterior distribution across each of the treatment combinations in the formula. *marginal_effects()* only gives summary statistics for the median, standard error, and credible intervals. To get the full distribution from which these summaries are calculated, I extracted iterations of fitted values for treatment x date interaction level using the the *fitted()* function. Within that, I specified "newdata" as simply the interaction levels from *marginal_effects()*. Here's the code.

```{r warning=FALSE,message=FALSE}
marg_m1_diet <- marginal_effects(model_diet_prior1, effects="species:date2", method="fitted")
marg_m1_full_post <- fitted(model_diet_prior1, newdata = marg_m1_diet$`species:date2`, summary = FALSE) #full_posterior from marginal_effects
columns <- paste(marg_m1_diet$`species:date2`$species,"_",marg_m1_diet$`species:date2`$date2)
marg_m1_full_post <- as.data.frame(marg_m1_full_post)
colnames(marg_m1_full_post) <- columns #full_posterior from marginal_effects
plot_marg1 <- gather(marg_m1_full_post, trt, total)
plot_marg1 <- plot_marg1 %>%
  separate(trt, c("species","date2"), sep=' _ ')


ggplot(plot_marg1, aes(x=total,y=species,fill=date2))+
  geom_density_ridges(alpha=0.3)+
  scale_x_log10()+
  scale_fill_grey()+
  theme(axis.title.y=element_blank())+
  xlab("Number of prey items per fish")+
  NULL



```

The figure above shows wide prior expectations. Perhaps wider than reasonable. For example, this model implies that there is a reasonable possibility that a fish could have more than 100,000 prey items in a single gut. That might be true for bacteria, but not for insects. Based on prior data from fish at the same sites (Warmbold 2016), spotfin shiner had a maximum of only 56 prey items in their guts. Let's adjust the prior and see if we can get a more reasonable prediction. The code below is the same as that above, but simply adjusts the priors to be more restrictive, as shown in the *prior* specification of the *brm()* statement.

Step 1: update the original model with a different prior statement. Note: you could just paste the full model and change the priors, but using *update()* sometimes speeds up the process, especially if the model doesn't need to be re-complied. 

```{r cache = T, warning = F, message = F, results = "hide"}
model_diet_priors2 <- update(model_diet_prior1, 
                             prior=c(prior(normal(2,2),class="Intercept"), #changed from 3,2 to 2,2
                                     prior(normal(0,2),class="b"))) #changed from 0,5 to 0,2
```

Step 2: Plot it.

```{r warning=FALSE,message=FALSE}
marg_m2_diet2 <- marginal_effects(model_diet_priors2, effects="species:date2", method="fitted")
marg_m2_full_post2 <- fitted(model_diet_priors2, newdata = marg_m2_diet2$`species:date2`, summary = FALSE) #full_posterior from marginal_effects
columns <- paste(marg_m2_diet2$`species:date2`$species,"_",marg_m2_diet2$`species:date2`$date2)
marg_m2_full_post2 <- as.data.frame(marg_m2_full_post2)
colnames(marg_m2_full_post2) <- columns #full_posterior from marginal_effects
plot_marg2 <- gather(marg_m2_full_post2, trt, total)
plot_marg2 <- plot_marg2 %>%
  separate(trt, c("species","date2"), sep=' _ ')


ggplot(plot_marg2, aes(x=total,y=species,fill=date2))+
  geom_density_ridges(alpha=0.3)+
  scale_x_log10()+
  scale_fill_grey()+
  theme(axis.title.y=element_blank())+
  xlab("Number of prey items per fish")+
  NULL



```

This seems much more reasonable. The x-axis only goes up to ~ 100,000, but nearly all of our prior expectation is contained below ~1,000. That is still a wide range, but is much better than before. 


#Fit the model
The model below now includes the data by simply removing the *sample_prior="only" statement from before. It also includes the updated priors.

```{r cache = T, warning = F, message = F, results = "hide"}
model_diet_1 <- brm(total ~ species*date2, data=subset(diet_abund2,species!="quillback"),
                         family=poisson(link="log"),
                         prior=c(prior(normal(3,2),class="Intercept"),
                         prior(normal(0,5),class="b")),
                         chains=4, iter=1000)  
```

Check the output summary

```{r}
print(model_diet_1)
```
The Rhat's look fine. At this point, I follow the advice of McElreath (2015) and try to avoid interpreting much from the numbers. That's much easier to do with plotting. Next let's simulate 10 extra dataset and plot them next to the original data to see if our model can reproduce data that look somewhat like the original. If it can't, then we've misspecified something.

```{r}
pp_check(model_diet_1,type="boxplot")
```

Simulated data is OK, but they also miss some of the outliers in the original data. Check the plots of posteriors over raw data. 

```{r warning=FALSE, message=FALSE}
model_diet1 <- model_diet_1
marg_m1_diet <- marginal_effects(model_diet1, effects="species:date2")
marg_m1_full_post <- fitted(model_diet1, newdata = marg_m1_diet$`species:date2`, summary = FALSE) #full_posterior from marginal_effects
columns <- paste(marg_m1_diet$`species:date2`$species,"_",marg_m1_diet$`species:date2`$date2)
marg_m1_full_post <- as.data.frame(marg_m1_full_post)

colnames(marg_m1_full_post) <- columns #full_posterior from marginal_effects

plot_marg1 <- gather(marg_m1_full_post, trt, total)
plot_marg1 <- plot_marg1 %>%
  separate(trt, c("species","date2"), sep=' _ ')
plot_marg1 <- as.data.frame(plot_marg1)

ggplot()+
  geom_boxplot(data=plot_marg1,aes(x=date2,y=total,group=interaction(date2,species),fill=species),
               position=position_dodge(width=.5),outlier.shape=NA,width=.4)+
  geom_point(data=subset(diet_abund2,species!="quillback"), aes(x=date2,
             y=total_1,group=species,shape=method),alpha=0.2,
             position=position_dodge(width=0.5))+
    scale_shape_manual(values=c(1,2))+
  scale_color_brewer(direction=-1)+
  scale_fill_brewer(direction=-1)+
  theme_classic()+
  theme(axis.title.x=element_blank())+
  ylab("Number of prey items in fish guts")+
  coord_cartesian(ylim=c(1,200))+
  ggtitle('Fish diets across sample dates')+
  NULL

```

This plot shows that bluegill have more prey items than other species, regardless of the date. The first date is also something to note. On June 13, all diets were estimated by sacrificing the fish and dissecting their guts. In the other two dates, we used gastric lavage on anaesthetized fish, which were then returned to the backwater. For that reason, any differences betwen June 13 or other dates could also be due either to differences in fish feeding or to differences our method. It is reassuring that the relative order of fish remains roughly the same, with bluegill always having more diet items than largemouth, which has more than River Shiner. Spotfin have the largest shift among dates, though that is likely due to the single fish at the top of the graph, which had 200 prey items. Of those 200, 196 were zooplankton. No other fish had much zooplankton.

#What about just insects in fish diets?
To see how the analysis changes if we include only insects, let's re-run the model above, but this time with total insects as the response, rather than total prey items.

First, we'll make a new column that sums across only columns that are insects.
```{r}
colnames(diet_abund2)
diet_abund2 <- diet_abund2 %>%
  mutate(tot_insect = corixid+ chiro_l+ chiro_p+ chiro_a+ mayfly+ unk_ins+ unk_adult_ins+
                                    coleo_ad+ coleo_l+terr_hymen+terr_hemip) 
head(diet_abund2)
```


```{r cache = T, warning = F, message = F, results = "hide"}
model_diet_insects <- brm(tot_insect ~ species*date2, data=subset(diet_abund2,species!="quillback"),
                         family=poisson(link="log"),
                         prior=c(prior(normal(3,2),class="Intercept"),
                         prior(normal(0,5),class="b")),
                         chains=4, iter=1000) 
```



Check the model

```{r}
print(model_diet_insects)
```
```{r}
pp_check(model_diet_insects,type="boxplot")
```

This is a little better than the previous one, though sill misses the outliers. But it's bounded at zero and has reasonable medians and spread.


Now re-plot the model over the raw data.

```{r}
marg_m1_insects <- marginal_effects(model_diet_insects, effects="species:date2")
marg_m1_insects_post <- fitted(model_diet_insects, newdata = marg_m1_insects$`species:date2`, summary = FALSE) #full_posterior from marginal_effects
columns <- paste(marg_m1_insects$`species:date2`$species,"_",marg_m1_insects$`species:date2`$date2)
marg_m1_insects_post <- as.data.frame(marg_m1_insects_post)

colnames(marg_m1_insects_post) <- columns #full_posterior from marginal_effects

plot_marg1_ins <- gather(marg_m1_insects_post, trt, total)
plot_marg1_ins <- plot_marg1_ins %>%
  separate(trt, c("species","date2"), sep=' _ ')
plot_marg1_ins <- as.data.frame(plot_marg1_ins)

ggplot()+
  geom_boxplot(data=plot_marg1_ins,aes(x=date2,y=total,group=interaction(date2,species),fill=species),
               position=position_dodge(width=.5),outlier.shape=NA,width=.4)+
  geom_point(data=subset(diet_abund2,species!="quillback"), aes(x=date2,                                                             y=tot_insect,group=species,shape=method),alpha=0.2,
             position=position_dodge(width=0.5))+
  scale_shape_manual(values=c(1,2))+
  scale_color_brewer(direction=-1)+
  scale_fill_brewer(direction=-1)+
  theme_classic()+
  theme(axis.title.x=element_blank())+
  ylab("Number of insects in fish guts")+
  ggtitle('Fish diets across sample dates - Insects only')+
  NULL
```

The major change with this model is that the estimate for spotfin on June 13 is now much lower, becuase we've removed the datapoint that indicated 198 zooplankton. As a result, it's clear that the relative number of prey items in fish guts is similar among species for each date. As mentioned above, this is also re-assuring that our method of gastric lavage captured a decent proportion of prey items in fish guts.

The data are summarized in the table below. Bluegill had an average of 43, 39, and 22 insects in their guts on each successive date. Largemouth bass had 15, 8, and 10. Spotfin had 14, 8, and 7.
```{r}
tbl_diet_total <- plot_marg1_ins %>%
  group_by(species,date2)%>%
  summarise(median = median(total),
            sd = sd(total),
            low_95 = quantile(total, probs=0.025),
            high_95 = quantile(total, probs=0.975))%>%
  mutate_if(is.numeric, round, 1)
tbl_diet_total
write.csv(tbl_diet_total, file="tbl_diet_total")
```

##Summary of diet totals
Based on this analysis, it seems OK to move ahead with to analyzing the stage-structure of fish feeding. The results from the gut dissections and gastric lavage are similar for abundance. Moreover, for analyzing proportions, such as the proportion of pupae in diets, any biases in total abundance are less important.



##Stage-structured feeding in fishes
This analysis will quantify the proportion of non-larval insects in each fishes gut. 

The first step is to add a variable that contains the total number of life-stages for each fish. However, we may not have life-stage information on each taxon. First, we tested for the dominant taxa by plotting the taxon proportions of each insect. This analysis is limited to insects because they are most likely to show distinct stages and are the only prey taxon that emerges out of the water, forming an aquatic-terrestrial linkage that is the focus of this study.
```{r insects in diets}
colnames(diet_abund2)

diet_abund2 %>%
  mutate(chiro_tot = chiro_l + chiro_p + chiro_a) %>%
  select(id,loc,method,date2,species,length_mm,corixid,mayfly,unk_ins,unk_adult_ins,sciaridae,coleo_ad,coleo_l,terr_hymen,chiro_tot,terr_hemip,total,tot_insect) %>%
  gather(taxon,number,"corixid":"terr_hemip")%>%
  filter(species!="quillback")%>%
  
  ggplot(aes(x=reorder(taxon,-number),y=number,color=date2))+
  geom_point(position=position_dodge(width=0.5))+
  theme_classic()+
  facet_wrap(~species)+
  scale_color_grey()+
  xlab("")+
  theme(axis.text.x = element_text(angle=90,hjust=1,vjust=0.5))+
  scale_y_log10()+
  ggtitle("Insect abundance in fish guts")+
  NULL

```
In the figure above, each data point is a diet sample (via dissection or gastric lavage) from an individual fish. These data clearly show that corixids and chironomids were the two most abundant insect prey items on each date. In fact, no fish contained more than two individuals of any taxon besides Corixidae and Chironomidae. Becuase chironomids were the most abundant insects in both our benthic and emergence samples, and because their life-stages (larval, pupal, adult) are relatively easy to identify in gut contents, we estimated stage-structured feeding in fishes using chironomids only.

```{r Chironomid stages in fish diets}
colnames(diet_abund2)

diet_chiros <- diet_abund2 %>%
  mutate(chiro_tot = chiro_l + chiro_p + chiro_a) %>%
  select(id,loc,method,date2,species,length_mm,chiro_tot,chiro_l,chiro_p,chiro_a) %>%
  gather(stage,number,"chiro_l":"chiro_a")%>%
  filter(species!="quillback")

ggplot(data=diet_chiros,aes(x=reorder(stage,-number),y=number,color=date2))+
  geom_point(position=position_dodge(width=0.5))+
  theme_classic()+
  facet_wrap(~species)+
  scale_color_grey()+
  xlab("")+
  theme(axis.text.x = element_text(angle=90,hjust=1,vjust=0.5))+
  scale_y_log10()+
  ggtitle("Insect abundance in fish guts")+
  NULL 
```

All fish had multiple larval and pupal stages, while adult stages were rare. Only three fish had an adult chironomid in their gut, with a total of four adults eaten out of a total of 424 chironomids.

#Proportion of non-larval chironomids in fish diets
Based on the summary above, we analyzed stage-structured feeding by measuring the proportion of non-larval chironomids in fish diets across dates.

```{r Prior model for proportion of chironomid stages in fishes, cache = T, warning = F, message = F, results = "hide"}
mod_stage <- brm(number|trials(chiro_tot)~species*date2, data=subset(diet_chiros,stage=="chiro_l"),
                 family=binomial(link="logit"),
                 prior=c(prior(normal(0.5,2),class="Intercept"),
                         prior(normal(0,1),class="b")),
                 chains=1,iter=500,sample_prior="only")
```
```{r}
marg_prior_stage <- marginal_effects(mod_stage, effects="species:date2")
marg_prior_stage2 <- fitted(mod_stage, newdata = marg_prior_stage$`species:date2`, summary = FALSE) #full_posterior from marginal_effects
columns <- paste(marg_prior_stage$`species:date2`$species,"_",marg_prior_stage$`species:date2`$date2)
marg_prior_stage_post <- as.data.frame(marg_prior_stage2)

colnames(marg_prior_stage_post) <- columns #full_posterior from marginal_effects

plot_marg_prior_stage <- gather(marg_prior_stage_post, trt, prop_larv)
plot_marg_prior_stage <- plot_marg_prior_stage %>%
  separate(trt, c("species","date2"), sep=' _ ')
plot_marg_prior_stage <- as.data.frame(plot_marg_prior_stage)
plot_marg_prior_stage$date2 <- factor(plot_marg_prior_stage$date2,levels=c("2017-06-28","2017-06-23","2017-06-13"))

ggplot()+
  geom_density_ridges(data=plot_marg_prior_stage,aes(x=1-prop_larv,y=date2,fill=species),alpha=0.5)+
  #geom_rug(data=subset(diet_chiros,stage=="chiro_l"), aes(y=date2,                                                             x=1-(number/chiro_tot),group=species,color=species))+
  scale_shape_manual(values=c(1,2))+
  scale_color_brewer(direction=-1)+
  scale_fill_brewer(direction=-1)+
  theme_classic()+
  theme(axis.title.y=element_blank())+
  ggtitle('Stage-structured feeding - prior predictive')+
  xlab("Proportion of non-larval chironomids in fish diets")+
  NULL
```

Becuase we don't know the proportion of pupae in diets, and it is not well-studied, this seems like a reasonable starting point, with the credible intervals spanning nearly the entire range.

Now we fit the model.

```{r Full model for proportion of chironomid stages in fishes, cache = T, warning = F, message = F, results = "hide"}
mod_stage_full <- brm(number|trials(chiro_tot)~species*date2, data=subset(diet_chiros,stage=="chiro_l"),
                 family=binomial(link="logit"),
                 prior=c(prior(normal(0.5,2),class="Intercept"),
                         prior(normal(0,1),class="b")))
```

```{r}
print(mod_stage_full)
```

```{r fig.width=5}
marg_stage <- marginal_effects(mod_stage_full, effects="species:date2")
marg_stage2 <- fitted(mod_stage_full, newdata = marg_stage$`species:date2`, summary = FALSE) #full_posterior from marginal_effects
columns <- paste(marg_stage$`species:date2`$species,"_",marg_stage$`species:date2`$date2)
marg_stage_post <- as.data.frame(marg_stage2)

colnames(marg_stage_post) <- columns #full_posterior from marginal_effects

plot_marg_stage <- gather(marg_stage_post, trt, prop_larv)
plot_marg_stage <- plot_marg_stage %>%
  separate(trt, c("species","date2"), sep=' _ ')
plot_marg_stage <- as.data.frame(plot_marg_stage)
plot_marg_stage$date2 <- factor(plot_marg_stage$date2,levels=c("2017-06-28","2017-06-23","2017-06-13"))

ggplot()+
  geom_density_ridges(data=plot_marg_stage,aes(x=1-prop_larv,y=date2,fill=species),alpha=0.5)+
  #geom_rug(data=subset(diet_chiros,stage=="chiro_l"), aes(y=date2,                                                             x=1-(number/chiro_tot),group=species,color=species))+
  scale_shape_manual(values=c(1,2))+
  scale_color_brewer(direction=-1)+
  scale_fill_brewer(direction=-1)+
  theme_classic()+
  theme(axis.title.y=element_blank())+
  ggtitle('Stage-structured feeding')+
  xlab("Proportion of non-larval chironomids in fish diets")+
  NULL
```

As shown in the figure, chironomid larvae were the dominant life-stage in the diets of all fishes in early June. Over the next 10 and 15 days, the proportion of pupal or adult (mostly pupal) stages increased for all fish, particularly bluegill and spotfin, but the increas was far greater for spotfin, which fed more heavily on pupal chironomids as emergence progressed. 

Here's a table summarizing the data in the plot above.
```{r}
tbl_diet_prop <- plot_marg_stage %>%
  group_by(species,date2) %>%
  summarize(median = median(1-prop_larv),
            sd = sd(1-prop_larv),
            low_95 = quantile(1-prop_larv, probs=0.025),
            high_95 = quantile(1-prop_larv, probs=0.975)) %>%
  arrange(species,desc(date2)) %>%
  mutate_if(is.numeric, round, 2)
tbl_diet_prop
write.csv(tbl_diet_prop, file="tbl_diet_prop.csv")
```

Finally, we can plot the prior versus posterior distribution for each treatment combination to see what we've learned from the data.
```{r}
plot_marg_prior_stage$distribution <- "prior"
plot_marg_stage$distribution <- "posterior"


prior_v_post <- rbind(plot_marg_stage,plot_marg_prior_stage)
colnames(prior_v_post)
prior_v_post$distribution <- factor(prior_v_post$distribution,levels=c("prior","posterior"))

ggplot(prior_v_post, aes(x=1-prop_larv, y=date2,fill=distribution))+
  geom_density_ridges(alpha=.9)+
  facet_wrap(~species)+
  scale_fill_brewer(direction=1)+
  theme_classic()+
  NULL
```


